---
title: "Gov 50 Final Project"
author: "Emory Paul"
description: "Final Project for GOV 50"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to homelessness, particularly as it relates to the experiences of different subsets of the unhoused population, such as chronically homeless individuals, veterans, and unaccompanied youth. For example, I'd love to investigate city-level data that would allow me to determine whether different strategies of service provision impact the magnitude of each subset of the unhoused population, though I am not committed to this idea. Overall, any data set related to homelessness or housing insecurity would be very interesting to explore given that these are my primary academic and professional interests. 

## Research Proposal

Does the criminalization of homelessness contribute to lower rates of unsheltered homelessness in urban communities? In this study, I plan to analyze the extent to which the community-level adoption of laws criminalizing homelessness through “quality of life” ordinances decrease rates of unsheltered homelessness. I hypothesize that the criminalization of homelessness does not significantly decrease rates of unsheltered homelessness in urban areas, thus demonstrating the futility of these punitive laws. Recently, there has been a proliferation of ordinances that criminalize the survival tactics of people experiencing homelessness in cities across the country. These laws, motivated by anti-homeless sentiment, funnel the unhoused into the homelessness-jail cycle, only further exacerbating their experience of homelessness. My sample is comprised of an aggregation of community-level data points for the over 400 Continuum of Cares (CoCs) across the United States, and I plan to filter these CoCs based on the "major_city" indicator variable in order to develop a sample of CoCs in urban communities. My unit of analysis is a Continuum of Care (CoC), which is a local planning body responsible for coordinating funding for homeless services in each community. The explanatory variables of interest are the number of municipal ordinances criminalizing homelessness in each community, which I plan to measure through the count of sleeping, camping, lying/sitting, and vehicle restriction laws (hou_pol_hlessconduct_sleep), count of loitering and vagrancy laws (hou_pol_hlessconduct_loiter), total count of prohibited conduct laws (hou_pol_hlessconduct_total), count of begging laws (hou_pol_hlessconduct_beg), and count of food sha=ring laws (hou_pol_hlessconduct_food). The outcome variable is the rate of unsheltered homeless per 10,000 people, which is coded as “pit_unshelt_pit_hud_share.” This variable is measured as the proportion of unsheltered homeless per 10,000 people in each CoC. If I do not observe a significant effect or if I observe a substantial increase in unsheltered homelessness over time in communities where there has been an increase in laws criminalizing homelessness, this would provide support for my hypothesis. However, if I observe a significant decrease in unsheltered homelessness over time in communities where there has been an increase in laws criminalizing homelessness, this would provide evidence against my hypothesis. When I run a regression of the laws criminalizing homelessness on the rates of unsheltered homelessness variable, a near-zero/zero or a positive, significant coefficient would demonstrate support for my hypothesis.

## First Data Visualization

```{r}
library(tidyverse)
homelessdata <- read_csv("/Users/emorypaul/Downloads/05b_analysis_file_update_v1.csv", 
                             show_col_types = FALSE)

homelessdata |>
  filter(cocnumber %in% c("TX-500", 
                          "TX-503", 
                          "TX-600", 
                          "TX-601", 
                          "TX-603", 
                          "TX-604", 
                          "TX-607", 
                          "TX-611", 
                          "TX-624", 
                          "TX-700", 
                          "TX-701"), 
         year == 2015) |>
  ggplot(aes(x = hou_pol_hlessconduct_sleep, 
             y = pit_tot_unshelt_pit_hud, 
             fill = cocnumber)) +
  geom_col() +
  labs(x = "Count of Sleeping, Camping, Lying/Sitting, and Vehicle Restriction Laws",
       y = "Total Unsheltered Homeless Pop.",
       title = "Outcome of Ordinances Criminalizing Homelessness")
```

For my first data visualization, I subset the data to solely the Continuum of Cares (CoC) in Texas, a state notorious for laws criminalizing homelessness. I also subset it to the year of 2015. I then plotted the count of sleeping, camping, lying/sitting, and vehicle restriction laws in each CoC against the total unsheltered homeless population in each of those CoCs to visualize their relationship. Upon first glance, it appears that there does not seem to be a significant relationship between the variables, which is in line with my hypothesis. However, I will need to continue exploring the data and accounting for confounding variables to be able to draw firm conclusions.

## Results from One Analysis

```{r}
homelessdata <- homelessdata |>
  filter(econ_urb_urbanicity == 3 | econ_urb_urbanicity == 4) |>
  mutate(criminalization_laws = hou_pol_hlessconduct_total + 
                                hou_pol_hlessconduct_sleep + 
                                hou_pol_hlessconduct_loiter + 
                                hou_pol_hlessconduct_beg +
                                hou_pol_hlessconduct_food) |>
  ggplot(aes(x = criminalization_laws, 
             y = pit_tot_hless_pit_hud)) + 
  geom_point() + 
  geom_smooth()

homelessdata
```

Largely Urban and Major City CoCs --> N = 832 . My question focuses on urban communities. 
Does the criminalization of homelessness contribute to lower rates of unsheltered homelessness in urban communities?
I hypothesize that the criminalization of homelessness does not significantly decrease rates of unsheltered homelessness in urban areas, thus demonstrating the futility of these punitive laws.
