---
title: "Gov 50 Final Project"
author: "Emory Paul"
description: "Final Project for GOV 50"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

With the increase of unsheltered homelessness in communities across the nation, policy makers, service providers, and members of the public have to begun to call for real, lasting solutions to this ever-expanding humanitarian crisis. While some elected officials have responded to these calls with benevolent policies of supportive services and affordable housing, others have responded with punitive laws criminalizing the survival tactics of people experiencing homelessness, such as sleeping outside and receiving food from passersby. Recent research has detailed how these laws, motivated by anti-homeless sentiment, funnel the unhoused into the homelessness-jail cycle, further exacerbating their experience of homelessness (Bauman). With this understanding in mind, I set out to investigate this issue with the following research question: What is the relationship between laws criminalizing homelessness and the amount of unsheltered homelessness in urban communities? In this study, I plan to analyze the relationship between the community-level adoption of laws criminalizing homelessness through “quality of life” ordinances and the extent of homelessness in those communities, which will allow me to better understand the effectiveness or lack thereof of these laws. I hypothesize that the criminalization of homelessness is associated with higher levels of homelessness in urban areas, pointing to the harmful nature of criminalization. The null hypothesis, in this case, would be that there is no relationship between the two variables. 

## The Data

My sample consists of an aggregation of community-level data points for the over 400 Continuum of Cares (CoCs) across the United States. CoCs, the unit of analysis for this study, are the local planning bodies responsible for coordinating funding for homeless services in each community. In the remainder of this study, I will use the term "community" and "CoC" interchangeably.

The key dependent variable of interest is the total amount of unsheltered homelessness in each urban CoC, which is coded as “pit_tot_unshelt_pit_hud." This variable is collected through in-person surveys on a single night in January, hence this study's cross-sectional design. The key independent variable of interest is the total number of  municipal ordinances criminalizing homelessness in each CoC, which I measure through the count of sleeping, camping, lying/sitting, and vehicle restriction laws (hou_pol_hlessconduct_sleep), count of loitering and vagrancy laws (hou_pol_hlessconduct_loiter), total count of prohibited conduct laws (hou_pol_hlessconduct_total), count of begging laws (hou_pol_hlessconduct_beg), and count of food sha=ring laws (hou_pol_hlessconduct_food). These data points were collected through an analysis of each community's "quality of life" ordinances by the National Law Center on Homelessness and Poverty (NLCHP). Additionally, to identify possible confounders, I measured the relationship between median gross rent for renter-occupied housing units and total unsheltered homelessness in each community. This variable was collected through surveys by the Census Bureau. 

I chose to focus on urban communities given that they, on average, have larger amounts of unsheltered homelessness as well as broader availability of data on this population than rural communities, which provides for a better study. Similarly, I chose to focus on unsheltered homelessness given that laws criminalizing homelessness target those sleeping outside, as opposed to those sleeping in shelters, given their heightened visibility. 

If I observe a positive relationship between total number of laws criminalizing homelessness and amount of unsheltered homelessness in urban communities, this would provide support for my hypothesis. However, if I observe a negative relationship or if the null hypothesis is confirmed through the lack of any relationship, this would provide evidence against my hypothesis. When I run a regression of the two variables, a positive coefficient with a p-value less than an alpha of 0.05 would demonstrate support for my hypothesis.


```{r}
library(tidyverse)
library(broom)
homelessdata <- read_csv("/Users/emorypaul/Downloads/05b_analysis_file_update_v1.csv", 
                             show_col_types = FALSE)
```

## Filtering and Mutating the Data 

The original data set lacks an aggregate variable of all the "quality of life" ordinances in each community, so I mutated the data to include a new variable, `criminalization_laws`, which is the total sum of all "quality of life" ordinances criminalizing

Given my study's focus on urban communities, I created a subset of the data named `homelessdata_urban` that only includes Continuum of Cares (CoCs) in "Largely Urban" or "Major City" areas. Within this same subset, I also filtered the data to exclude CoCs which had a total unsheltered homeless population above 20,000 since these communities are outliers that skewed my past plots. 


```{r}
homelessdata <- homelessdata |>
  mutate(criminalization_laws = hou_pol_hlessconduct_total + 
                                hou_pol_hlessconduct_sleep + 
                                hou_pol_hlessconduct_loiter + 
                                hou_pol_hlessconduct_beg +
                                hou_pol_hlessconduct_food)

homelessdata_urban <- homelessdata |>
  filter(econ_urb_urbanicity == 3 | econ_urb_urbanicity == 4) |>
  filter(pit_tot_unshelt_pit_hud < 20000)
```


## Summary of the Dependent Variable

As a summary of the main dependent variable in my study, I plotted the total amount unsheltered homelessness in all CoCs over time (2010-2017). The below bar plot indicates that since 2014, unsheltered homelessness has been increasing in communities across the United States. With this study, I hope to better understand whther the criminalization of homelessness has contributed to this increase. 

```{r}
summary_homeless <- homelessdata_urban |>
  ggplot(aes(x = year, 
             y = pit_tot_unshelt_pit_hud)) +
  geom_col(fill = "aquamarine4") +
  labs(x = "Year",
       y = "Total Unsheltered Homeless Population",
       title = "Total Unsheltered Homeless Population Over Time (Figure 1)")

summary_homeless
```


## Results

## Main Regression

The following output is a regression equation using my primary independent variable, total number of laws criminalizing homelessness. 

```{r}
fit_main <- lm(pit_tot_unshelt_pit_hud ~ criminalization_laws, data = homelessdata_urban)

var_labels <- c("(Intercept)" = "Intercept",
                "criminalization_laws" = "Laws Criminalizing Homelessness")

modelsummary::modelsummary(fit_main,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

As seen in the above table, the intercept has a coefficient of 130.241, which can be interpreted as the following: when a community does not have any laws criminalizing homelessness, the CoC has around 130 individuals experiencing unsheltered homelessness within its boundaries. Furthermore, the variable measuring the criminalization of homelessness in each community has a coefficient of 61.086, which can be interpreted as the following: each additional law criminalizing homelessness is correlated with an increase in a community's total unsheltered homeless population by around 61 persons. Given that the p-value for this coefficient, which is less than 0.001, is smaller than the pre-specified significance level, alpha = 0.05, the coefficient is statistically significant, thus supporting my hypothesis that there is a positive relationship between the criminalization of homelessness and the amount of homelessness in urban communities. However, this relationship can not be determined as causal given the potential presence of confounding variables.  

## Plot of Main Regression

The following graph is a scatter plot displaying the above regression using the main independent variable, total number of laws criminalizing homelessness, plotted against . 

```{r}
homeless_criminalization <- homelessdata_urban |>
  ggplot(aes(x = criminalization_laws, 
             y = pit_tot_unshelt_pit_hud)) + 
  geom_point(color = "aquamarine4") + 
  geom_smooth(method = "lm", se = FALSE, color = "brown3", size = 1.5) +
   labs(x = "Total Number of Laws Criminalizing Homelessness",
       y = "Total Unsheltered Homeless Population",
       title = "Criminalization of Homelessness in Urban CoCs (Figure 2)")

homeless_criminalization
```

As the plot shows, there is a positive correlation between criminalization of homelessness and the amount of homelessness in urban communities. 

## Multiple Regression

Though my main regression demonstrates a strong relationship between laws criminalizing homelessness and amount of unsheltered homelessness in communities, it is important to consider potential confounding variables that may be skewing the relationship. Given that homelessness is fundamentally a housing problem, I decided to run another regression equation that includes the median gross rent of renter-occupied housing units in an effort to see the relationship between housing prices and amount of unsheltered homelessness in each community. The regression is displayed in the below output. 

```{r}
fit_multiple <- lm(pit_tot_unshelt_pit_hud ~ criminalization_laws + hou_mkt_medrentgross_acs5yr,
                   data = homelessdata_urban)

var_labels <- c("(Intercept)" = "Intercept",
                "criminalization_laws" = "Laws Criminalizing Homelessness",
                "hou_mkt_medrentgross_acs5yr" = "Median Gross Rent (Renter-Occupied Housing Units)")

modelsummary::modelsummary(fit_multiple,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

As seen in the above table, the intercept has a coefficient of −1687.31, though this intercept is not plausible given that it is not impossible to have a negative amount of unsheltered homeless individuals. Nonetheless, the variable measuring the confounding variable of median gross rent has a coefficient of 2.089, which can be interpreted as the following: each additional dollar in median gross rent of renter-occupied units is correlated with an increase in a community's total unsheltered homeless population by around 2 persons. Moreover, the variable measuring the criminalization of homelessness in each community has a coefficient of 56.642, which can be interpreted as the following: each additional law criminalizing homelessness is correlated with an increase in a community's total unsheltered homeless population by around 56 persons. Because this value is lower than the one calculated in the previous regression that lacked the confounding variable, it is possible that median gross rent in communities could be contributing to the positive relationship between criminalization and unsheltered homelessness, hence why a causal relationship cannot be determined. The p-values for both of these coefficients, which are less than 0.001, are smaller than the pre-specified significance level, alpha = 0.05, so the coefficients are statistically significant. 

## Plot of Multiple Regression

The following graph is a scatter plot displaying the above regression using the confounding variable, median gross rent of renter-occupied housing units. 


```{r}
homeless_rentprice <- homelessdata_urban |>
  ggplot(aes(x =  hou_mkt_medrentgross_acs5yr, 
             y = pit_tot_unshelt_pit_hud)) + 
  geom_point(color = "aquamarine4") + 
  geom_smooth(method = "lm", se = FALSE, color = "brown3", size = 1.5) +
  labs(x = "Median Gross Rent (Renter-Occupied Housing Units)",
       y = "Total Unsheltered Homeless Population",
       title = "Total Unsheltered Homeless Pop. vs. Median Gross Rent (Figure 3)")

homeless_rentprice
```

The above scatter plot confirms the regression output that there is a positive correlation between median gross rent and the amount of unsheltered homelessness in urban communities. Therefore, housing prices are a potential confounding variable. 


## Conclusion

Overall, there appears to be a positive, statistically significant relationship between the criminalization of homelessness through "quality of life" ordinances and the amount of unsheltered homelessness in urban communities. These results support my hypothesis that the criminalization of homelessness through "quality of life" ordinances is correlated with higher levels of homelessness in urban CoCs. The results also support the rejection of the null hypothesis of no relationship. This observed relationship suggests that the criminalization of homelessness could only be exacerbating the extent of homelessness in communities, though the relationship can not be labelled as causal, so more research is needed to whether the laws criminalizing homeless are a cause or an effect of high amounts of unsheltered homelessness. Furthermore, I identified that median gross rent of renter-occupied units is likely a confounding variable influencing the relationship between criminalization and total unsheltered homeless population. However, there are several limitations of my study. For example, though the Point-In-Time (PIT) count that I used for the measurement of unsheltered homelessness is the most reliable measure of the data point that exists, but it is still a mere rough estimate of the total unsheltered population given just how difficult it is to count this population. Furthermore, I used total unsheltered homelessness as opposed to unsheltered homeless per capita because I lacked sufficient data on the matter, which limits th Finally, the data set that I used only went up to 2017, which limits the external validity of the study, though my results do still indicate trends that we are likely still seeing today. To address these limitations and explore the observed relationship in greater depth, future research should control for population size as well as incorporate  other confounding variables, such as housing prices, eviction rates, and median income, into the analysis. 

## Works Cited

Bauman, Tristia. “Housing Not Handcuffs: Ending the Criminalization of Homelessness in U.S. Cities.” National Homelessness Law Center, homelesslaw.org/wp-content/uploads/2018/10/Housing-Not-Handcuffs.pdf. 

Nisar, Hiren, et al. “Market Predictors of Homelessness: How Housing and Community Factors Shape Homelessness Rates within Continuums of Care.” HUD: Office of Policy Development and Research, 2019, www.huduser.gov/portal/publications/Market-Predictors-of-Homelessness.html.

Nisar, Hiren, et al. "Homeless Prediction Model.” HUD: Office of Policy Development and Research, 2019, https://www.huduser.gov/portal/datasets/hpmd.html.


