---
title: "Gov 50 Final Project"
author: "Emory Paul"
description: "Final Project for GOV 50"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction


## The Data

```{r}
library(tidyverse)
library(broom)
homelessdata <- read_csv("/Users/emorypaul/Downloads/05b_analysis_file_update_v1.csv", 
                             show_col_types = FALSE)
```

## Filtering and Mutating the Data 

```{r}
homelessdata <- homelessdata |>
  filter(econ_urb_urbanicity == 3 | econ_urb_urbanicity == 4) |>
  mutate(criminalization_laws = hou_pol_hlessconduct_total + 
                                hou_pol_hlessconduct_sleep + 
                                hou_pol_hlessconduct_loiter + 
                                hou_pol_hlessconduct_beg +
                                hou_pol_hlessconduct_food)
```




## Summary of the Dependent Variable

```{r}
homelessdata |>
  ggplot(data = general_homeless, aes(x = year, 
             y = pit_unshelt_pit_hud_share)) +
  geom_col()
```




## Results

```{r}
 homeless_criminalization <- homelessdata |>
  ggplot(aes(x = criminalization_laws, 
             y = pit_tot_unshelt_pit_hud)) + 
  geom_point(color = "steelblue") + 
  geom_smooth(method = "lm", se = FALSE, color = "indianred1", size = 1.5) +
   labs(x = "Total Number of Laws Criminalizing Homelessness",
       y = "Total Unsheltered Homeless Population",
       title = "Total Unsheltered Homeless Pop. vs. Criminalization in Urban CoCs")

homeless_criminalization

fit <- lm(pit_tot_unshelt_pit_hud ~ criminalization_laws, data = total_homeless_criminalization) 
fit2 <- lm(pit_tot_unshelt_pit_hud ~ criminalization_laws + total_rent_inventory_acs5yr, data = total_homeless_criminalization) 

fit
fit2
```


## Results from  Analysis

```{r}
percapita_homeless_criminalization <- homelessdata |>
  filter(econ_urb_urbanicity == 3 | econ_urb_urbanicity == 4) |>
  mutate(criminalization_laws = hou_pol_hlessconduct_total + 
                                hou_pol_hlessconduct_sleep + 
                                hou_pol_hlessconduct_loiter + 
                                hou_pol_hlessconduct_beg +
                                hou_pol_hlessconduct_food)

  ggplot(data = percapita_homeless_criminalization, aes(x = criminalization_laws, 
             y = pit_unshelt_pit_hud_share)) + 
  geom_point(color = "steelblue") + 
  geom_smooth(method = "lm", se = FALSE, color = "indianred1", size = 1.5) +
   labs(x = "Total Number of Laws Criminalizing Homelessness",
       y = "Total Homeless Population",
       title = "Total Homeless Population vs. Criminalization of Homelessness in Urban Areas")

fit3 <- lm(pit_unshelt_pit_hud_share ~ criminalization_laws, data = percapita_homeless_criminalization) 
fit4 <- lm(pit_unshelt_pit_hud_share ~ criminalization_laws + total_rent_inventory_acs5yr, data = percapita_homeless_criminalization) 

fit3
fit4
```



```{r}
percapita_homeless_criminalization <- homelessdata |>
  filter(econ_urb_urbanicity == 3 | econ_urb_urbanicity == 4) |>
  mutate(criminalization_laws = hou_pol_hlessconduct_total + 
                                hou_pol_hlessconduct_sleep + 
                                hou_pol_hlessconduct_loiter + 
                                hou_pol_hlessconduct_beg +
                                hou_pol_hlessconduct_food)

  ggplot(data = percapita_homeless_criminalization, aes(x = econ_labor_incineq_acs5yr, 
             y = pit_tot_unshelt_pit_hud)) +
    geom_point() + 
    geom_smooth()
```

## Conclusion


## Works Cited

Nisar, Hiren, et al. “Market Predictors of Homelessness: How Housing and Community Factors Shape Homelessness Rates within Continuums of Care.” HUD: Office of Policy Development and Research, 2019, www.huduser.gov/portal/publications/Market-Predictors-of-Homelessness.html. 



